---
title: 'Week 2 Assignment : Course Project 1'
author: "Ike"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
header-includes: 
  - \usepackage{color}
  - \usepackage{framed}
fig_Caption: true

---
----------------------------------------

```{r global, include=TRUE, warning=FALSE, message=FALSE, highlight=TRUE, fig.dim=c(6,5), fig.align='center'}
# 
knitr::opts_chunk$set(echo = TRUE, comment = NA)
library(DT)
library(xtable)
library(dplyr)
library(ggplot2)
library(plotly)
#convert date to a Date variable
mydata <- data.frame(read.csv("activity.csv", stringsAsFactors = FALSE))
mydata$date <- as.Date(mydata$date)

```
---------------------------------------


## Introduction

This assignment used data collected each day, at five minute intervals over a 2 month period 
with a personal activity monitoring device. 

---------------------------------------
  
## Part 1

#### The mean, sum and, median number of steps taken per day

Here, missing values in the **steps** column were excluded. The rest of the data was grouped by date
and the daily mean, total number of steps and, median values for each day were returned 
These values are shown below as column values of summarized data for each day. The total number of
observations are **`r nrow(mydata) `**


------------------------------------------
 
```{r showtable, results='asis', warning=FALSE, message=FALSE}

# Remove all observations with missing values
mydata <- na.omit(mydata)
#
    DataSummary <- mydata %>% 
	      group_by(date) %>% 
	      summarize( MeanSteps = round(mean(steps), 2), 
			            SumSteps = sum(steps), 
			            MedianSteps = median(steps))
knitr::kable(head(DataSummary, 10), caption = ("The mean, total number of steps and median steps per day"))


```
--------------------------------------------

#### Histogram of total number of steps taken each day


```{r, warning=FALSE, message=FALSE}

gplot1 <- ggplot(data=DataSummary, aes(SumSteps)) + 
    geom_histogram(alpha = .5, col="blue", 
                 aes(fill=..count..)) + 
      labs(title="Histogram of Total Number of Steps Per Day",
       x = "Total number of steps taken each day")
return(ggplotly(gplot1))


```

#### Average Daily Activity Pattern  


Below, is a time series plot of daily activities with a vertical line was drawn at five minute interval
with maximum number of steps.


```{r}
#
maxint <- max(mydata$steps)
gplot2 <- ggplot(data=mydata, aes(x=interval, y=steps)) + 
  		    geom_point() +
		      geom_vline(xintercept = maxint, linetype="dotted", 
                color = "red", size=1.5) +
          labs(title="Time series plot of steps vs intervals",
          x = "Intervals")
return(ggplotly(gplot2))

```

#### The five minute interval with maximum number of steps across the data set is `r maxint `

------------------------------------------------

## Part 2


#### Activity data with missing values imputed


```{r}

#Load data again and count the number of observations with missing values

mydataM <- data.frame(read.csv("activity.csv", stringsAsFactors = FALSE))
mydataM$date <- as.Date(mydataM$date)
missingData <- sum(is.na(mydataM))


```
--------------------------------------------

#### The number of observations with missing values is  `r missingData ` 

The mising values are only in the **steps** taken column. We will replace each missing value with the mean
value of that column and verify that new dataset has zero missing values afterwards.

```{r}

mydataM$steps[is.na(mydataM$steps)] <- mean(mydataM$steps, na.rm=TRUE)
missingDataM <- sum(is.na(mydataM))


```

#### After imputing mising values, the number of observations with missing values is  `r missingDataM ` 
The total number of observations with missing values replaced is **`r nrow(mydataM) `**


----------------------------------------------

#### The mean, sum and, median number of steps each day with missing values imputed



```{r, warning=FALSE, message=FALSE}
    DataSum <- mydataM %>% 
	      group_by(date) %>% 
	      summarize(M_Steps = round(mean(steps), 2), 
			            S_Steps = round(sum(steps), 2),
			            MedSteps = round(median(steps), 2))
knitr::kable(head(DataSum, 10), caption = ("The Mean, Total and Median steps per day"))
#
gplot3 <- ggplot(data=DataSum, aes(M_Steps)) + 
  geom_histogram(alpha = .5, col="blue", 
                 aes(fill=..count..)) + 
    labs(title="Histogram of the total steps without missing values imputed",
       x = "Total number of steps taken each day")

return(ggplotly(gplot3))


```

The results show that replacing all 2304 missing values with mean values has the effect of increasing
overall values (overall mean, total number of steps and, median) compared to previous values with missing 
observation removed.

------------------------------

## Part 3



#### Possible differences in activity pattern during weekdays and weekends

```{r}
   mydataM$date <- as.Date(mydataM$date)
#create a factor variable 'WDays' of weekdays and weekends
WeekDays <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
mydataM$WDays <- factor((weekdays(mydataM$date) %in% WeekDays), 
         levels=c(FALSE, TRUE), labels=c('Weekend', 'Weekday') )
#
table(mydataM$WDays)

```


---------------------------------------------

#### A time series of data for weekdays and weekends 

```{r}

    ggplot(data = mydataM) +
	     geom_point( mapping = aes(x=interval, y = steps) ) +
	              facet_wrap(~WDays) 


```

---------------------------------------------

#### Filter data for weekdays and weekends

```{r}
WeekEnd  <- mydataM %>% 
	 filter(WDays == 'Weekend') 
#
     ggplot(data = WeekEnd) +
         geom_line( mapping = aes(x=interval, y = steps) ) +
       labs(title="WeekEnd")

WeekDay  <- mydataM %>% 
		group_by(date, interval, WDays) %>%
        filter(WDays == 'Weekday')  
#
 ggplot(data = WeekDay) +
    geom_line( mapping = aes(x=interval, y = steps) ) +
   labs(title="WeekDay")

```


